---
title: "3. Explore Advanced Analyses"
author: "Qingzhou Zhang"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Advanced Analysis and Interoperability with ComplexMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
```

# Introduction

While the `ComplexMap::createComplexMap()` function provides a seamless, one-step workflow for most use cases, the package also includes a suite of granular functions for users who require more control, wish to benchmark their results, or need to export data for other tools.

This vignette covers these advanced topics:

1.  **Granular Workflow Control**: Running the quality control and refinement steps manually to inspect intermediate results.

2.  **Benchmarking Predictions**: Using `evaluateComplexes()` to compare your experimental complexes against a reference set like CORUM.

3.  **Advanced Network Exploration**: Adding theme information back to a `ComplexMap` object to enable powerful, theme-based queries.

4.  **Interoperability**: Exporting the final network for visualization and analysis in external software like Cytoscape.

We will use the package's built-in datasets for this demonstration.

```{r load-data}
# Load the experimental and reference complex lists
utils::data("demoComplexes", package = "ComplexMap")
utils::data("referenceComplexes", package = "ComplexMap")

```

# 1. Granular Workflow Control

The `createComplexMap()` function automatically runs quality control and refinement. However, you can run these steps manually using the underlying functions to better understand and tune your analysis.

### 1.1 Quality Control

First, we can run `qcComplexList()` on our raw input data to get a detailed report on complex sizes and redundancy.

```{r qc, message=TRUE}
# Run the QC function with verbose output
ComplexMap::qcComplexList(demoComplexes)
```

The report provides summary statistics and warns us that a number of complex pairs are highly redundant (Jaccard similarity >= 0.8). This confirms that the refinement step is necessary.

### 1.2 Manual Refinement

Next, we can manually run `refineComplexList()`. This allows us to inspect the set of complexes *after* merging but *before* functional enrichment, which can be useful for debugging or parameter tuning.

```{r refine}
message("Number of complexes before refinement: ", length(demoComplexes))

# Merge complexes with a Jaccard similarity of 0.75 or higher
refinedComplexes <- ComplexMap::refineComplexList(
  demoComplexes,
  mergeThreshold = 0.75,
  verbose = FALSE # Set to TRUE for detailed messages
)

message("Number of complexes after refinement: ", length(refinedComplexes))
```
Running this step manually gives us direct access to the `refinedComplexes` list for further inspection before proceeding with the rest of the workflow.

# 2. Benchmarking Predictions

For methods development, it is essential to evaluate how well a set of predicted complexes matches a "gold standard" reference. The `evaluateComplexes()` function calculates four standard metrics for this purpose.

We will compare our `refinedComplexes` against the `referenceComplexes` (a curated subset of CORUM).

```{r evaluate}
# Evaluate the refined complexes against the reference set
evaluationMetrics <- ComplexMap::evaluateComplexes(
  predictedComplexes = refinedComplexes,
  referenceComplexes = referenceComplexes,
  nCores = 2,
  verbose = FALSE
)

# Print the resulting metrics
print(evaluationMetrics)
```

*   **PPV** (Positive Predictive Value): Measures the accuracy of your predictions.

*   **Sn** (Sensitivity): Measures the coverage of the reference set.

*   **Acc** (Accuracy): The geometric mean of PPV and Sn.

*   **MMR** (Maximum Matching Ratio): A score based on finding the best one-to-one mapping between predicted and reference complexes.

These metrics provide a quantitative assessment of your complex prediction quality.

# 3. Advanced Network Exploration: Querying by Theme

In the main workflow vignette, we used `summarizeThemes()` to get a table of biological themes. A powerful advanced technique is to add this theme information back into the `ComplexMap` object itself, enabling direct queries.

First, let's generate a `ComplexMap` object.
```{r create-map-again}
gmtPath <- ComplexMap::getExampleGmt()
gmt <- ComplexMap::getGmtFromFile(gmtPath, verbose = FALSE)

# We use the manually refined list as input
cm_obj <- ComplexMap::createComplexMap(
  complexList = refinedComplexes,
  gmt = gmt,
  verbose = FALSE
)
```

Now, we generate the themes and add the theme assignments to our node table.
```{r add-themes}
# Generate the theme summary to get theme labels
theme_summary <- ComplexMap::summarizeThemes(cm_obj, verbose = FALSE)

# To get the theme assignment for each node, we must re-run community detection
nodes <- ComplexMap::getNodeTable(cm_obj)
edges <- ComplexMap::getEdgeTable(cm_obj)
graph <- igraph::graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)
communities <- igraph::cluster_louvain(graph)

# Create a mapping from themeId (integer) to themeLabel (character)
theme_map <- theme_summary[, c("themeId", "themeLabel")]

# Add theme ID and label columns to our node table
nodes$themeId <- igraph::membership(communities)
nodes <- dplyr::left_join(nodes, theme_map, by = "themeId")

# **Crucially, update the object with the new node information**
cm_obj$nodes <- nodes
```

With the prepared object, we can now find all complexes belonging to a specific theme, for example, the "26S Proteasome".

```{r query-theme}
# To make the example robust, we query for a theme label that we know
# exists because we just generated it in the `theme_summary` table.
query_label <- theme_summary$themeLabel

message("Dynamically querying for a theme found in the map: ", query_label)

theme_complexes <- ComplexMap::queryMap(
  cm_obj,
  query = query_label,
  type = "theme"
)

# Show the results
theme_complexes %>%
  dplyr::select(complexId, themeLabel, proteinCount, betweenness)
```

This technique provides a powerful way to programmatically subset and analyze the biological modules within your map.

# 4. Interoperability: Exporting to Cytoscape

While `ComplexMap` provides excellent internal visualizations, you may want to use other powerful tools like [Cytoscape](https://cytoscape.org/) for advanced analysis, network manipulation, or creating publication-quality figures.

The `exportNetwork()` function writes the node and edge tables to disk as tab-separated files (`.tsv`), which can be easily imported into Cytoscape.

```{r export}
# Use a temporary directory for this example
temp_dir <- tempdir()
file_prefix <- file.path(temp_dir, "human_complex_map")

# Export the network
ComplexMap::exportNetwork(cm_obj, filePrefix = file_prefix)

# List the files that were created
list.files(temp_dir, pattern = "human_complex_map")
```

To use these in Cytoscape, you would:

1.  Go to `File > Import > Network from File...` and select the `_edges.tsv` file.

2.  In the mapping dialog, map the `source_complex_id` column to "Source Node" and `target_complex_id` to "Target Node".

3.  Once the network is loaded, go to `File > Import > Table from File...` and select the `_nodes.tsv` file.

4.  Ensure the "Key column for network" is set to `complexId`.

This will import all your node attributes (colors, sizes, functional domains, etc.) and map them directly onto your network.

# Conclusion

This vignette has demonstrated the advanced capabilities of the `ComplexMap` package. Beyond the simple one-step workflow, the package provides granular functions for quality control, rigorous benchmarking tools for methods development, and powerful utilities for deep, theme-based querying and exporting results to other platforms. This layered functionality makes `ComplexMap` a comprehensive toolkit for both novice and expert users.