---
title: "3. Advanced Analysis and Visualization"
author: "Qingzhou Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Advanced Analysis and Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE
)
```

## Advanced Analysis Workflow

Once you have a `ComplexMap` object, the package provides several functions for deeper analysis and interpretation of the network structure. This vignette covers:

1.  **Optimizing Parameters**: Using `evaluateComplexes()` to tune the `refineComplexList` step.
2.  **Summarizing & Querying Themes**: A streamlined workflow using `summarizeThemes()` and `queryMap()` to identify and investigate biological clusters.

```{r setup}
library(ComplexMap)
library(dplyr)
```

### 1. Optimizing Refinement with `evaluateComplexes`

The `mergeThreshold` in `refineComplexList` determines how aggressively redundant complexes are merged. The default is a conservative **Jaccard** similarity of 0.90 to preserve biological variants.

The `evaluateComplexes()` function can score a refined complex list against a gold standard, helping you tune this parameter.

```{r evaluate}
# Load the package's demo and reference data
data(demoComplexes)
data(referenceComplexes)

# 1. Run Refinement
refinement_output <- refineComplexList(
  demoComplexes,
  mergeThreshold = 0.90,
  similarityMethod = "jaccard",
  verbose = FALSE
)

# 2. Evaluate
# Pass the $refinedComplexes element to the evaluation function
metrics <- evaluateComplexes(
  predictedComplexes = refinement_output$refinedComplexes,
  referenceComplexes = referenceComplexes,
  nCores = 2, # Use multiple cores if available
  verbose = FALSE
)

# Display the resulting metrics
knitr::kable(as.data.frame(metrics))
```

For more systematic tuning, see the documentation for `benchmarkParameters()`, which automates this process over a range of thresholds.

### 2. A Streamlined Workflow for Theme Analysis

A key goal of `ComplexMap` is to organize complexes into a functional landscape. `summarizeThemes()` uses community detection to find densely connected physical neighborhoods (themes) and, by default, adds this information directly to your `ComplexMap` object. This makes querying by theme simple and intuitive.

First, let's generate a map object for this example.

```{r create_adv_map}
gmtPath <- getExampleGmt()
gmt <- getGmtFromFile(gmtPath, verbose = FALSE)

cm_obj_adv <- createComplexMap(
  complexList = demoComplexes,
  gmt = gmt,
  mergeThreshold = 0.85, 
  verbose = FALSE
)
```

#### Step 2.1: Summarize Themes and Add to Object

Now, we run `summarizeThemes()`. The default `add_to_object = TRUE` returns the modified `ComplexMap` object, now containing theme information in its node table.

```{r summarize_and_add}
# Ensure reproducibility of community detection
set.seed(123)

# This single command finds themes AND adds them to the object
cm_obj_adv <- summarizeThemes(cm_obj_adv, verbose = FALSE)

# We can inspect the new columns in the node table
getNodeTable(cm_obj_adv) %>%
  select(complexId, primaryFunctionalDomain, themeId, themeLabel) %>%
  head() %>%
  knitr::kable()
```

The returned object is now "theme-aware".

#### Step 2.2: Querying the Map for Specific Information

The `queryMap()` function can now be used with `type = "theme"` to easily retrieve all complexes belonging to a specific biological theme.

**Query by Protein**

Find all complexes that contain a specific protein.

```{r query_protein}
# Find all complexes containing the protein "UBA1"
queryMap(cm_obj_adv, query = "UBA1", type = "protein") %>%
  select(complexId, primaryFunctionalDomain, proteinCount, themeLabel, proteins) %>%
  knitr::kable()
```

**Query by Complex ID**

Retrieve the data for a single complex.

```{r query_complex}
# Get the data for complex "CpxMap_0001"
queryMap(cm_obj_adv, query = "CpxMap_0001", type = "complex") %>%
  # Use glimpse to see all columns, including theme info
  glimpse()
```

**Query by Theme**

Because we added theme labels to the object, we can now query them directly.

```{r query-theme}
# First, let's find a valid theme label to query from our map
node_table <- getNodeTable(cm_obj_adv)
if (nrow(node_table) > 0 && "themeLabel" %in% names(node_table)) {
  target_theme <- na.omit(node_table$themeLabel)

  cat("Querying for all nodes in theme:", target_theme, "\n\n")

  # Query the map using the theme label
  theme_nodes <- queryMap(cm_obj_adv, query = target_theme, type = "theme")
  
  theme_nodes %>%
    select(complexId, primaryFunctionalDomain, themeLabel) %>%
    head() %>%
    knitr::kable()
}
```

This streamlined workflow makes exploring the thematic structure of the complex map efficient and powerful.